---
title: "NZaidi_LiveSessionAssigment_W2"
author: "Najeeb Zaidi"
date: "August 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Requirements:

Use the stackloss data to create the following:

     * Use summary() function on data set and interpret the results.
     * Use the ???str??? function on the data set and interpret the results.
     * Create a scatter plot of stack.loss vs. Air.Flow and describe the relationship.
     * Copy and paste the final (error-free) Rmarkdown code into the window provided at 2DS section 2.8.3.

### 1. Upload / Read Data:
```{r}
# reading R package builtin data set
library(datasets)
data(stackloss)

summary(stackloss)
```

### 2A. View Structure of StackLoss dataset:
```{r}
str(stackloss)

typeof(stackloss$Air.Flow)
typeof(stackloss$Water.Temp)
typeof(stackloss$Acid.Conc.)
typeof(stackloss$stack.loss)

head(stackloss)

```

### 2B. Interpreting the above results:

**"stackloss"** is a data frame with 21 observations and 4 variables. The type of each vector within this data frame is also listed in output above. 

The Header information lists the first 6 records stored in this dataframe.
Summary function presented earlier describes the statistical characteristics of data stored in this dataframe.


### 3A. Scatterplot of stack.loss vs. Air.Flow

```{r}
library(ggplot2)
scatter.smooth(stackloss$Air.Flow, stackloss$stack.loss)

cor(stackloss$Air.Flow, stackloss$stack.loss)

```

### 3B. Interpretation of Scatterplot

The dataframe suggests that stackloss is a dependent variable of Air Flow, Water Temperature and Acid Concentration levels. This implies that a regression relationship may exist.

**The scatter plot where Air.Flow is taken as 'x' or independent variable while stack.loss is the 'y' or dependent variable describes a positive linear relationship between these two.**

Additionally, the cor() function computes the correlation between them.

We can further run a linear regression with lm() function and find correation as ANOVA to evaluate the extent and nature of Air Flow impacting the stack.loss.

From [background reading](http://www.calvin.edu/~stob/courses/m241/F11/outlines/Nov20-multilinear.pdf), we get to know that stackloss is dependent variable of a multiple regression where the measurements were collected from an ammonia oxidation plant. Source link provided to view full details.










Najeeb Zaidi - MSDS 6306
.
